---
title: "Manual MAIT cell annotation in Lett dataset using AUCell package and Garner et al gene set intersection with Zheng et al"
subtitle: "scRNA-seq workflow series part4" 
title-block-banner: "#f0f3f5"
title-block-banner-color: "black"
author: "Ivan Osinnii"
date: "18 April 2024"
format: html
code-fold: true
editor: visual
toc: true
toc-location: left
number-sections: true
fig-cap-location: bottom
css: style.css
---
# Introduction



# Load data and dependencies

```{r, warning= FALSE, message= FALSE, echo= TRUE}
getwd()
library(SingleCellExperiment)
library(SummarizedExperiment)
library(RSQLite)
library(AUCell)
library(ggplot2)
library(RColorBrewer)
library(tidyverse)
library(readxl)
library(scater)
library(uwot)
library(GSEABase)
```

### Load previously processed Lett dataset, check how cell look like on UMAP

```{r, warning= FALSE, message= FALSE, echo= TRUE, fig.height= 6, fig.width= 8}
getwd()

lett.sce <- readRDS("./../input/Lett.sce.liver.gene.symbol_analyzed.rds")

lett.sce$MonacoImmuneData <- lett.sce@colData@listData[["cellReferences_celldex_MonacoImmuneData"]]@listData[["labels"]]

plotUMAP(lett.sce, colour_by = 'MonacoImmuneData',point_size = 1.0)
```

### Load the list of MAIT marker genes

```{r}
genes <-read_excel("./../input/CellMarkerMAITcells.xlsx", sheet = 2, col_names = TRUE) ## Garner list intersection with Zheng list

```

# Analysis

### Transform MAIT marker list into GeneSet object using GSEABase functions

```{r}
all.sets <- lapply(names(genes), function(x) {
  GeneSet(genes[[x]], setName=x)        
})
all.sets <- GeneSetCollection(all.sets)
```

### Now model cell type rankings using AUCell package function

```{r}
rankings <- AUCell_buildRankings(counts(lett.sce),
                                 plotStats=FALSE, verbose=FALSE)
### Calculate AUC curves for each cell taking our markers
cell.aucs <- AUCell_calcAUC(all.sets, rankings)
results <- t(assay(cell.aucs))
str(results)
```

### compare new cell projection and Monnaco Immune Data 

```{r}
new.labels <- colnames(results)[max.col(results)]
tab <- table(new.labels, lett.sce$MonacoImmuneData)
tab

```
We can see that few other cell types likely got mislabeled as MAIT cells

### Assign our new labels to colData of lett.sce

```{r}

class(new.labels)
new_labels_df <- DataFrame(new.labels)
class(new_labels_df)
# Now assign the DataFrame to colData
colData(lett.sce) <- new_labels_df
lett.sce$AUCell_cells <- lett.sce$new.labels

```

### Let's plot newly assigned MAIT labels on UMAP

```{r, fig.height= 6, fig.width= 8}
plotUMAP(lett.sce, colour_by = "AUCell_cells", point_size = 1.0)

```

### Now we calculate the amount of MAIT cells and their proportion to all T cells

```{r}
MAIT_cell_values <- colData(lett.sce)$AUCell_cells 
value_counts <- table(MAIT_cell_values)
value_counts
percent3 <- round((value_counts[1] /4375 * 100),1)
```

We output these results 

```{r}
message("Number of MAIT cells equals ", value_counts[1])
message("Number of all T cells equals ", 4375)
message("Persentage of MAIT cells out of all T cells found with this method of annotation equals ", percent3)

```

### Conclusion