---
title: "Manual MAIT cell annotation in Lett dataset (blood) using AUCell package and Garner et al gene set intersection with Zheng et al"
subtitle: "scRNA-seq workflow series part 6" 
title-block-banner: "#f0f3f5"
title-block-banner-color: "black"
author: "Ivan Osinnii"
date: "07 May 2024"
format: html
code-fold: true
editor: visual
toc: true
toc-location: left
number-sections: true
fig-cap-location: bottom
css: style.css
---

# Introduction

# Load data and dependencies

```{r, warning= FALSE, message= FALSE, echo= TRUE}
getwd()
library(SingleCellExperiment)
library(SummarizedExperiment)
library(RSQLite)
library(AUCell)
library(ggplot2)
library(RColorBrewer)
library(tidyverse)
library(readxl)
library(scater)
library(uwot)
library(GSEABase)
```

### Load previously processed Lett dataset, check how cell look like on UMAP

```{r, warning= FALSE, message= FALSE, echo= TRUE, fig.height= 6, fig.width= 8}
getwd()

lett.sce <- readRDS("./../input/Gil_analyzed.rds")

lett.sce$MonacoImmuneData <- lett.sce@colData@listData[["cellReferences_celldex_MonacoImmuneData"]]@listData[["labels"]]

plotUMAP(lett.sce, colour_by = 'MonacoImmuneData',point_size = 1.0)
```

### Load the list of MAIT marker genes

```{r}
genes <-read_excel("./../input/CellMarkerMAITcells.xlsx", sheet = 2, col_names = TRUE) ## Garner list intersection with Zheng list

```

# Analysis

### Transform MAIT marker list into GeneSet object using GSEABase functions

```{r}
all.sets <- lapply(names(genes), function(x) {
  GeneSet(genes[[x]], setName=x)        
})
all.sets <- GeneSetCollection(all.sets)
```

### Now model cell type rankings using AUCell package function

```{r}
rankings <- AUCell_buildRankings(counts(lett.sce),
                                 plotStats=FALSE, verbose=FALSE)
### Calculate AUC curves for each cell taking our markers
cell.aucs <- AUCell_calcAUC(all.sets, rankings)
results <- t(assay(cell.aucs))
str(results)
```

### compare new cell projection and Monnaco Immune Data

```{r}
new.labels <- colnames(results)[max.col(results)]
tab <- table(new.labels, lett.sce$MonacoImmuneData)
tab

```

We can see that few other cell types likely got mislabeled as MAIT cells

### Assign our new labels to colData of lett.sce

```{r}

class(new.labels)
new_labels_df <- DataFrame(new.labels)
class(new_labels_df)
# Now assign the DataFrame to colData
colData(lett.sce) <- new_labels_df
lett.sce$AUCell_cells <- lett.sce$new.labels

```

### Let's plot newly assigned MAIT labels on UMAP

```{r, fig.height= 6, fig.width= 8}
plotUMAP(lett.sce, colour_by = "AUCell_cells", point_size = 1.0)

```

### Now we calculate the amount of MAIT cells and their proportion to all T cells

```{r}
MAIT_cell_values <- colData(lett.sce)$AUCell_cells 
value_counts <- table(MAIT_cell_values)
value_counts
percent3 <- round((value_counts[1] /4333 * 100),1)
```

We put these results out

```{r}
message("Number of MAIT cells equals ", value_counts[1])
message("Number of all T cells equals ", 4333)
message("Persentage of MAIT cells out of all T cells found with this method of annotation equals ", percent3)

```

### Conclusion

Methodologically the results of blood cell analysis are partially similar to those of liver analysis. We get the highest percentage of MAIT cells by using transcriotional signature taken from Garner et al. dataset and AUCell package. But in contrast to liver when we use SingleR (automated) cell annotation, we get less MAIT cells with MonacoImmuneData and more with Garner et al. label projection. Aditionally, it looks like we significantly enriched our blood sample for MAIT cells with magnetic bead selection, Because the percentage of MAIT cells is unsusually high as for blood sample.

Session Information

</button>

::: {#collapseOne .accordion-collapse .collapse}
<div>

```{r}
sessionInfo()
```

</div>
:::